On the Synchronization of Computational RFIDs

### 2018-09-07 ieee-tmc-latex/article.tex
image licence

\begin{figure}
	\centering
	\includegraphics[width=0.70\columnwidth]{figures/scenario.png}
	\caption{\label{fig:motivation}A representative greenhouse monitoring application using CRFIDs. Two readers are deployed to provide power to the individual tags and to collect their periodic sensor readings.}
\end{figure}

@@

\myitem{Frequent power failures:} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g. the clock offset and relative frequency, into the non-volatile memory to recover when they harvested sufficient energy to start up. However, saving computational state is also an energy consuming task~\cite{blisp}.

@@

As presented in Fig.~\ref{fig:crfid_testbed}, a typical CRFID system is composed of an RFID reader, a single WISP tag placed inside the communication range of this reader and a host computer to control this reader. In our testbed, we used a 915\,MHz Impinj Speedway R1000 RFID reader with firmware version 3.2.4 connected to a Laird S9028PCR 8.5\,dBic gain antenna. We placed the WISP tag at the line-of-sight from the reader antenna. For the host-reader control operations, we used sllurp~\cite{sllrp_github}, a LLRP (Low-Level Reader Protocol)~\cite{llrp}  control library written in Python.  To program the WISP tag, we used MSP430 Flash Emulation Tool (FET), in combination with TI Code Composer Studio (CCS), attached to the host.  All of our experimental evaluation presented in the next sections were conducted using this setup at a university office with human presence.

@@

In this section, we introduce a simple but novel \emph{event-based} reader-tag synchronization inspired by simple PI controllers and provide its advantages over the previous approach. In event-based synchronization, a common event which is observable by all receiver devices simultaneously is generated by the reference device. Upon receiving events generated at regular intervals, receiver devices can predict the occurrence time of future events. In order to synchronize the tag with such a mechanism, the RFID reader does not send explicit timestamp values as in sender-receiver based synchronization but instead generates events at regular intervals. Upon observing these events, the receiver tag adjusts the rate of its software clock so that it predicts the occurrence of the next event precisely. 
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/Event-Based.png}
	\caption{\label{fig:event_based}Event-Based synchronization steps: The tag timestamps successive \emph{BlockWrite} events and adjusts its software clock.}
\end{figure}
We explored the EPC Gen2 standard to see how to generate events at regular intervals and realized that the \emph{BlockWrite} operation allows this feature.\footnote{It is possible to exploit modulation properties of signals from the reader to tags (such as bit timing in FM0, i.e. Miller 1, modulation) in order to get the timing measurements of the regular patterns in the signal. However, this requires using a high-precision clock to obtain microsecond-precision timing measurements. In this study, we used the higher-level properties of the EPC Gen2 standard, rather than exploiting its low-level signal properties. By means of this, we could  use the external 32\,KHz clock that runs even in low-power standby mode (see Section \ref{sec:Clock_Hardware}) to obtain timing measurements.)} Fig.~\ref{fig:event_based} presents the steps of event-based synchronization. During the command phase, the EPC Gen2 standard allows a maximum of eight successive \emph{BlockWrite} operations. It is desirable to use the first and the last \emph{BlockWrite} events for synchronization since it is better to compensate frequency differences observed in longer time intervals to adjust the software clock.

@@

The real-time length between the first and the last \emph{BlockWrite} operation is the \emph{event period}, we denote by $\tau$, and its variation is the main error source. Therefore, it is important to explore its characteristics. We sniffed the communication protocol between the RFID reader and a tag presented in Fig. \ref{fig:event_based}. We took 20 sample measurements about the \emph{event period}, which is summarized in Fig.~\ref{fig:block_write_delay}. According to our measurements, the \emph{event period} was distributed with a mean of 226.76\,ms and standard deviation of 0.41\,ms; respectively. We would like to mention that these values are dependent on the RFID Reader implementation of the EPC Gen2 protocol.

@@

Let $t_0$ denote the first event reception time at the tag. Using the pre-measured event period $\tau$ of the RFID reader, the tag can estimate the next event reception time at time $t_1$ as
\begin{eqnarray}
\hat{C}_w(t_1) &=& C_w(t_0)+\tau(1+\hat{f}_w^r(t_0)), \label{eq:core}
\end{eqnarray}
where $\hat{f}_w^r(t_0)$ denotes the \emph{estimated relative frequency} at time $t_0$, which is an estimate for \eqref{eq:relative_frequency}. The intuition behind \eqref{eq:core} is that $1+\hat{f}_w^r(t_0)$ gives the estimated value of $f_r(t_0)/f_w(t_0)$, $\tau f_r(t_0)$ gives the estimated local time passed on the reader side and $\tau f_r(t_0)/f_w(t_0)$ gives the estimated number of clock ticks on the tag side that will pass until the next packet reception. Redefining \eqref{eq:gamma_error}, the \emph{prediction error} at time $t_1$ can be written as
\begin{eqnarray}
\gamma(t_1) & \!\!\!\!\!=\!\!\!\!\!\! & 
C_{w}(t_{1}\!)\!-\!\hat{C}_{w}(t_{1}\!) \!=\! 
C_{w}(t_{1}\!)\!-\!C_{w}(t_{0}\!)\!-\!\tau(1\!\!+\!\!\hat{f}_{w}^{r}(t_0\!)). \label{eq:gamma_t_1}
\end{eqnarray}
Apparently, the objective is to minimize the prediction error so that the event reception time is estimated correctly. As can be observed from \eqref{eq:gamma_t_1}, the estimation error on the relative clock frequency at time $t_0$, i.e. $\hat{f}_w^r(t_0)$, depends on the mean event period $\tau$, and its variance (since they affect $C_w(t_1)-C_w(t_0)$). Inspired by the integral controller update rule in \cite[Section III, (5)]{pi2015}, we propose to update the estimated relative frequency at time $t_1$ as
\begin{eqnarray}
\hat{f}_{w}^{r}(t_1) &=& \hat{f}_{w}^{r}(t_0)+\beta\gamma(t_1), 
\label{eq:update_rule}
\end{eqnarray}
where $\beta$ is the \emph{integral gain}. Assume that the uncertainty of the event period is $\varepsilon\sim\mathcal{N}(0,\sigma^2_{\tau})$ and the instantaneous change on the relative clock frequency is $\frac{df_{w}^{r}(t)}{dt}\sim\mathcal{N}(0,\sigma_{\eta}^{2})$. Under these assumptions, it can be proven that if $0<\beta<\frac{2}{\tau}$ is satisfied the synchronization will hold eventually; i.e. $\hat{f}_{w}^{r}=f_{w}^{r}$. Moreover, the mean and the variance ($\sigma^2_\gamma$) of the steady-state (asymptotic) estimation error of WISP-Sync can be given by 
\begin{eqnarray}
\gamma & \!\!\!\!\!\sim\!\!\!\!\! & \mathcal{N} \!\!\left(\!\!0,\sigma^2_\gamma\!\!=\!\!\frac{\sigma^2_{\eta}\!\left(\!\tau^2\!-\!\beta\tau^3\!+\!\frac{\beta^2\tau^4}{3}\!\right)\!\!+\!\!\beta^2\sigma^2_{\tau}\tau}{(2\beta-\beta^2\tau)}\!\!+\!\!\frac{\sigma^2_{\eta}\tau^3}{3}\!+\!\sigma^2_{\tau} \!\!\right),
\label{eq:sync_performance}
\end{eqnarray}
which is derived in Section A.2 of the Appendix.

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/sinan.png}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
received his Ph.D. degree in Computer Engineering from Ege University, Izmir, Turkey, in 2012. He was a post-doc at the Department of Information Engineering, University of Padova in 2013 and at Embedded Software Group, Delft University of Technology between 2015 and 2017. Currently, he is an assistant professor at the Department of Computer Engineering, Ege University and also a research affiliate with the Embedded Software Group at Delft University of Technology. His research interests are embedded systems and software, distributed algorithms and wireless networking.
\end{IEEEbiography}

>>>>

\begin{figure}
	\centering
	\includegraphics[width=0.70\columnwidth]{figures/scenario.png}
	\caption{\label{fig:motivation}A representative greenhouse monitoring application using CRFIDs. Two readers are deployed to provide power to the individual tags and to collect their periodic sensor readings. (Image by Joi Ito, source~\cite{greenhouse-image} licensed under CC BY 2.0.) }
\end{figure}

@@

\myitem{Frequent power failures:} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss. Therefore, they need to save the synchronization state, e.g. the clock offset and relative frequency, into the non-volatile memory to recover when they harvest sufficient energy to start up. However, this is also an energy consuming task~\cite{blisp}.

@@

As presented in Fig.~\ref{fig:crfid_testbed}, a typical CRFID system is composed of an RFID reader, a single WISP tag placed inside the communication range of this reader and a host computer to control this reader. In our testbed, we used a 915\,MHz Impinj Speedway R1000 RFID reader with firmware version 3.2.4 connected to a Laird S9028PCR 8.5\,dBic gain antenna. We placed the WISP tag at the line-of-sight from the reader antenna. For the host-reader control operations, we used sllurp~\cite{sllrp_github}, a LLRP (Low-Level Reader Protocol)~\cite{llrp}  control library written in Python.  To program the WISP tag, we used MSP430 Flash Emulation Tool (FET), in combination with TI Code Composer Studio (CCS), attached to the host.  Our experimental evaluations presented hereafter were conducted using this setup at a university office with human presence.

@@

In this section, we introduce a simple but novel \emph{event-based} reader-tag synchronization inspired by simple PI controllers and provide its advantages over the previous approach. In event-based synchronization, a common event which is observable by all receiver devices simultaneously is generated by the reference device. Upon receiving events generated at regular intervals, receiver devices can predict the occurrence time of future events. In order to synchronize the tag with such a mechanism, the RFID reader does not send explicit timestamp values as in sender-receiver based synchronization but instead generates events at regular intervals. Upon observing these events, the receiver tag adjusts the rate of its software clock so that it predicts the occurrence of the next event precisely. 
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/Event-Based.png}
	\caption{\label{fig:event_based}Event-Based synchronization steps: The tag timestamps successive \emph{BlockWrite} events and adjusts its software clock.}
\end{figure}
We explored the EPC Gen2 standard to see how to generate events at regular intervals and realized that the \emph{BlockWrite} operation allows this feature.\footnote{It is possible to exploit modulation properties of signals from the reader to tags (such as bit timing in FM0, i.e. Miller 1, modulation) in order to get the timing measurements of the regular patterns in the signal. However, this requires using a high-precision clock to obtain microsecond-precision timing measurements. In this study, we used the higher-level properties of the EPC Gen2 standard, rather than exploiting its low-level signal properties. By means of this, we could  use the external 32\,KHz clock that runs even in low-power standby mode (see Section \ref{sec:Clock_Hardware}) to obtain timing measurements.)} Fig.~\ref{fig:event_based} presents the steps of event-based synchronization. During the command phase, the EPC Gen2 allows a maximum of eight successive \emph{BlockWrite} operations. It is desirable to use the first and the last \emph{BlockWrite} events for synchronization since it is better to compensate frequency differences observed in longer time intervals to adjust the software clock.

@@

The real-time length between the first and the last \emph{BlockWrite} operation is the \emph{event period}, we denote by $\tau$, and its variation is the main error source. Therefore, it is important to explore its characteristics. We sniffed the communication protocol between the RFID reader and a tag presented in Fig. \ref{fig:event_based}. We took 20 sample measurements about the \emph{event period}, which is summarized in Fig.~\ref{fig:block_write_delay}. According to our measurements, the \emph{event period} was distributed with a mean of 226.76\,ms and standard deviation of 0.41\,ms; respectively. We would like to mention that these values are dependent on the RFID Reader implementation of the EPC Gen2.

@@

Let $t_0$ denote the first event reception time at the tag. Using the pre-measured event period $\tau$ of the reader, the tag can estimate the next event reception time at time $t_1$ as
\begin{eqnarray}
\hat{C}_w(t_1) &=& C_w(t_0)+\tau(1+\hat{f}_w^r(t_0)), \label{eq:core}
\end{eqnarray}
where $\hat{f}_w^r(t_0)$ denotes the \emph{estimated relative frequency} at time $t_0$, which is an estimate for \eqref{eq:relative_frequency}. The intuition behind \eqref{eq:core} is that $1+\hat{f}_w^r(t_0)$ gives the estimated value of $f_r(t_0)/f_w(t_0)$, $\tau f_r(t_0)$ gives the estimated local time passed on the reader side and $\tau f_r(t_0)/f_w(t_0)$ gives the estimated number of clock ticks on the tag side that will pass until the next packet reception. Redefining \eqref{eq:gamma_error}, the \emph{prediction error} at time $t_1$ can be written as
\begin{eqnarray}
\gamma(t_1) & \!\!\!\!\!=\!\!\!\!\!\! & 
C_{w}(t_{1}\!)\!-\!\hat{C}_{w}(t_{1}\!) \!=\! 
C_{w}(t_{1}\!)\!-\!C_{w}(t_{0}\!)\!-\!\tau(1\!\!+\!\!\hat{f}_{w}^{r}(t_0\!)). \label{eq:gamma_t_1}
\end{eqnarray}
Apparently, the objective is to minimize the prediction error so that the event reception time is estimated correctly. As can be observed from \eqref{eq:gamma_t_1}, the estimation error on the relative clock frequency at time $t_0$, i.e. $\hat{f}_w^r(t_0)$, depends on the mean event period $\tau$, and its variance (since they affect $C_w(t_1)-C_w(t_0)$). Inspired by the integral controller update rule in \cite[Section III, (5)]{pi2015}, we propose to update the estimated relative frequency at time $t_1$ as
\begin{eqnarray}
\hat{f}_{w}^{r}(t_1) &=& \hat{f}_{w}^{r}(t_0)+\beta\gamma(t_1), 
\label{eq:update_rule}
\end{eqnarray}
where $\beta$ is the \emph{integral gain}. Assume that the uncertainty of the event period is $\varepsilon\sim\mathcal{N}(0,\sigma^2_{\tau})$ and the instantaneous change on the relative clock frequency is $\frac{df_{w}^{r}(t)}{dt}\sim\mathcal{N}(0,\sigma_{\eta}^{2})$. Under these assumptions, it can be proven that if $0<\beta<\frac{2}{\tau}$ is satisfied the synchronization will hold eventually; i.e. $\hat{f}_{w}^{r}=f_{w}^{r}$. Moreover, the mean and the variance ($\sigma^2_\gamma$) of the steady-state (asymptotic) estimation error of WISP-Sync can be given by 
\begin{eqnarray}
\gamma & \!\!\!\!\!\sim\!\!\!\!\! & \mathcal{N} \!\!\left(\!\!0,\sigma^2_\gamma\!\!=\!\!\frac{\sigma^2_{\eta}\!\left(\!\tau^2\!-\!\beta\tau^3\!+\!\frac{\beta^2\tau^4}{3}\!\right)\!\!+\!\!\beta^2\sigma^2_{\tau}\tau}{(2\beta-\beta^2\tau)}\!\!+\!\!\frac{\sigma^2_{\eta}\tau^3}{3}\!+\!\sigma^2_{\tau} \!\!\right),
\label{eq:sync_performance}
\end{eqnarray}
which is derived in Section A.2 of the Appendix.

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{author_photos/sinan.png}}]{Kas{\i}m Sinan Y{\i}ld{\i}r{\i}m}
received his Ph.D. degree in Computer Engineering from Ege University, Izmir, Turkey, in 2012. He was a post-doc at the Department of Information Engineering, University of Padova in 2013 and at Embedded Software Group, Delft University of Technology between 2015 and 2017. Currently, he is an assistant professor at the Department of Computer Engineering, Ege University and also a member of the Embedded Software Group at Delft University of Technology. His research interests are embedded systems, wireless networks and distributed algorithms.
\end{IEEEbiography}
